\LaTeXe\ guide for the Journal of Functional Programming

### 2019-10-19 papers/jfp/2-background.tex
Work on S1-3

\noindent
The above makefile lists three tasks: (i) compile a utility library comprising
files \cmd{util.h} and \cmd{util.c} into \cmd{util.o} by
executing\footnote{In this example we pretend \cmd{gcc} is a pure function for
  the sake of simplicity. In reality, there are multiple versions of \cmd{gcc}.
  To account for this, the actual binary for \cmd{gcc} is often listed as a dependency, along
  with any supporting binaries, or standard libraries (such as \cmd{stdio.h}),
  that are used by \cmd{gcc}.
} the command
\cmd{gcc -c util.c}, (ii) compile the main source file \cmd{main.c} into
\cmd{main.o}, and (iii) link object files \cmd{util.o} and \cmd{main.o} into the
executable \cmd{main.exe}. The makefile contains the complete information about
the \emph{task dependency graph}, which is shown in Fig.~\ref{fig-make}(a).

@@

\noindent
The following property is essential for build systems; indeed, it is their
raison d'\^etre:

@@

@@

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{text}
A1: 10      B1: A1 + A2
A2: 20
\end{minted}
\vspace{1mm}

@@

\vspace{-1mm}
\subsection{\Shake: Dynamic Dependencies with No Remorse}
\label{sec-background-shake}

\Shake was developed to solve the issue of dynamic
dependencies~\cite{mitchell2012shake} without sacrificing
the minimality requirement. Building on the \Make example
from~\S\ref{sec-background-make}, we add the following files whose
dependencies are shown in Fig.~\ref{fig-shake}(a):

@@

\begin{minted}[xleftmargin=10pt]{haskell}
"release.tar" %> \_ -> do
    need ["release.txt"]
    files <- lines <$> readFile "release.txt"
    need files
    cmd "tar" $ ["-cf", "release.tar"] ++ files
\end{minted}

@@

\Shake's implementation is different from both \Make and \Excel in two aspects.
First, to decide which files need to be rebuilt,
it uses the dependency graph that is dynamically constructed during
the previous build. This idea has a long history, going back to
\emph{incremental}~\cite{demers1981incremental},
\emph{adaptive}~\cite{acar2002adaptive}, as well as
\emph{self-adjusting computations} -- see
Acar~\etal~\shortcite{acar2007selfadjusting} and~\S\ref{sec-related}. Second,
instead of aborting and deferring the execution of tasks whose newly discovered
dependencies have not yet been built (as \Excel does), \Shake \emph{suspends}
their execution until the dependencies are brought up to date. We refer to this
task scheduling algorithm as \emph{suspending}, see a further discussion
in~\S\ref{sec-suspending} and a concrete implementation
in~\S\ref{sec-implementation-shake}.

@@

\Shake also supports the \emph{early cutoff optimisation}. When it
executes a task and the result is unchanged from the previous build, it is
unnecessary to execute the dependent tasks, and hence \Shake can stop a build
earlier, as illustrated in Fig.~\ref{fig-cutoff}. Not all build systems support
early cutoff: \Shake and \Bazel (introduced below) do, but \Make and \Excel do
not; see~\S\ref{sec-dirty-bit} for an explanation of why.

@@

\begin{figure}
% \vspace{-2mm}
\begin{subfigure}[b]{0.25\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/bazel-example-checkout.pdf}}
\vspace{-0.5mm}
\caption{Download sources}
\end{subfigure}
\begin{subfigure}[b]{0.40\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/bazel-example-build.pdf}}
\vspace{-0.5mm}
\caption{Build \cmd{main.exe}}
\end{subfigure}
\begin{subfigure}[b]{0.31\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/bazel-example-rebuild.pdf}}
\vspace{-0.5mm}
\caption{Modify \cmd{util.c} and rebuild}
\end{subfigure}
\vspace{-1mm}
\caption{A cloud build example: (a)~download sources, (b)~build \cmd{main.exe}
by downloading it from the cloud and skipping intermediate files (only their
hashes are needed), (c)~modify \cmd{util.c} and rebuild \cmd{main.exe}, which
requires building \cmd{util.o} (nobody has compiled \cmd{util.c} before) and
downloading \cmd{main.o} (it is needed for linking \cmd{main.exe}). File hashes
are shown in circles, and non-materialised intermediates in dashed rounded
rectangles.\label{fig-bazel}}
\vspace{-1mm}
\end{figure}

@@

\Bazel is one of the first openly-available cloud build systems.
As of writing, it is not possible to express dynamic dependencies in
user-defined build rules; however some of the pre-defined build rules require
dynamic dependencies and the internal build engine can cope with them by using
a \emph{restarting} task scheduler, which is similar to that of \Excel but does
not use the calc chain. \Bazel is not minimal in the sense that it may restart a
task multiple times as new dependencies are discovered and rebuilt, but it
supports the early cutoff optimisation.

To support cloud builds, \Bazel maintains (i)~a \emph{content-addressable cache}
that maps the hash of a file's content to the actual content of that file;
(ii)~a memo table that records all executed build commands
with their input and output file hashes.
The memo table allows the build engine to bypass the execution
of a task, by predicting the hash of the result from the hashes of its
dependencies; then the content-addressable cache allows the engine
to download the result (if needed) based on the result hash. A concrete
implementation is provided in~\S\ref{sec-implementation-cloud}.

@@

\begin{table}
% \vspace{-2mm}
\caption{Build system differences.\label{tab-summary}}
% \vspace{-4mm}
% \smaller
\centering
\begin{tabular}{llllccc}
\hline
Build system\gap & Persistent build information\gap & Scheduler\gap   & Dependencies\gap    & Minimal\gap & Cutoff\gap & Cloud\\\hline
\Make       & File modification times      & Topological\gap & Static          & Yes     & No     & No   \\
\Excel      & Dirty cells, calc chain      & Restarting  & Dynamic         & No      & No     & No   \\
\Shake      & Previous dependency graph    & Suspending  & Dynamic         & Yes     & Yes    & No   \\
\Bazel      & Cloud cache, command history\gap & Restarting  & Dynamic$^{(*)}$ & No      & Yes    & Yes  \\\hline
\end{tabular}
% \vspace{2mm}
$^{(*)}$At present, user-defined build rules cannot have dynamic dependencies.
% \vspace{-8mm}
\end{table}

>>>>

\noindent
The above makefile lists three tasks: (i) compile a utility library comprising
files \cmd{util.h} and \cmd{util.c} into \cmd{util.o} by
executing\footnote{In this example we pretend \cmd{gcc} is a pure function for
  the sake of simplicity. In reality, there are multiple versions of \cmd{gcc}.
  To account for this, the actual binary for \cmd{gcc} is often also listed as a
  dependency, along with any supporting binaries, or standard libraries (such as
  \cmd{stdio.h}), that are used by \cmd{gcc}.
} the command
\cmd{gcc -c util.c}, (ii) compile the main source file \cmd{main.c} into
\cmd{main.o}, and (iii) link object files \cmd{util.o} and \cmd{main.o} into the
executable \cmd{main.exe}. The makefile contains the complete information about
the \emph{task dependency graph}, which is shown in Fig.~\ref{fig-make}(a).

@@

The fewer tasks are executed in a partial rebuild, the better. To be more
specific, consider the following property, which is essential for build systems;
indeed, it is their raison d'\^etre:

@@

\noindent
This property is tightly linked to build system \emph{correctness}, which we
will be ready to define in~\S\ref{sec-build-correctness}; for now, we will use
minimality as the guiding principle when exploring the design space of build
systems.

@@

\vspace{0.5mm}
\begin{minted}[xleftmargin=10pt]{text}
A1: 10      B1: A1 + A2
A2: 20
\end{minted}
\vspace{0.5mm}

@@

\subsection{\Shake: Dynamic Dependencies with No Remorse}
\label{sec-background-shake}

\Shake was developed to solve the issue of dynamic
dependencies~\cite{mitchell2012shake} without sacrificing
the minimality requirement.

Building on the \Make example
from~\S\ref{sec-background-make}, we add the following files whose
dependencies are shown in Fig.~\ref{fig-shake}(a):

@@

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
"release.tar" %> \_ -> do
    need ["release.txt"]
    files <- lines <$> readFile "release.txt"
    need files
    cmd "tar" $ ["-cf", "release.tar"] ++ files
\end{minted}
\vspace{1mm}

@@

\Shake's implementation is different from both \Make and \Excel in two aspects.
First, to decide which files need to be rebuilt, it stores the \emph{dependency
graph} that is constructed during the previous build (\mbox{instead of
just file modification times or a linear chain}). This idea has a long history,
going back to \emph{incremental}~\cite{demers1981incremental},
\emph{adaptive}~\cite{acar2002adaptive}, and
\emph{self-adjusting computations}~--~see
Acar~\etal~\shortcite{acar2007selfadjusting} and~\S\ref{sec-related}. Second,
instead of aborting and deferring the execution of tasks whose newly discovered
dependencies have not yet been built (as \Excel does), \Shake \emph{suspends}
their execution until the dependencies are brought up to date. We refer to this
task scheduling algorithm as \emph{suspending}, see a further discussion
in~\S\ref{sec-suspending} and a concrete implementation
in~\S\ref{sec-implementation-shake}.

@@

\Shake also supports the \emph{early cutoff optimisation}, which is illustrated
in Fig.~\ref{fig-cutoff}. When it executes a task and the result is unchanged
from the previous build, it is unnecessary to execute the dependent tasks, and
hence \Shake can stop a build earlier. Not all build systems support early
cutoff: \Shake and \Bazel (introduced below) do, but \Make and \Excel do not;
see~\S\ref{sec-dirty-bit} for an explanation of why.

@@

\begin{figure}
% \vspace{-2mm}
\begin{subfigure}[b]{0.25\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/bazel-example-checkout.pdf}}
% \vspace{-0.5mm}
\caption{Download sources}
\end{subfigure}
\begin{subfigure}[b]{0.40\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/bazel-example-build.pdf}}
% \vspace{-0.5mm}
\caption{Build \cmd{main.exe}}
\end{subfigure}
\begin{subfigure}[b]{0.31\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/bazel-example-rebuild.pdf}}
% \vspace{-0.5mm}
\caption{Modify \cmd{util.c} and rebuild}
\end{subfigure}
% \vspace{-1mm}
\caption{A cloud build example: (a)~download sources, (b)~build \cmd{main.exe}
by downloading it from the cloud and skipping intermediate files (only their
hashes are needed), (c)~modify \cmd{util.c} and rebuild \cmd{main.exe}, which
requires building \cmd{util.o} (nobody has compiled \cmd{util.c} before) and
downloading \cmd{main.o} (it is needed for linking \cmd{main.exe}). File hashes
are shown in circles, and non-materialised intermediates in dashed rounded
rectangles.\label{fig-bazel}}
% \vspace{-1mm}
\end{figure}

@@

\Bazel is one of the first openly-available cloud build systems.
As of writing, it is not possible to express dynamic dependencies in
user-defined build rules; however some of the pre-defined build rules require
dynamic dependencies and the internal build engine can cope with them by using
a \emph{restarting} task scheduler, which is similar to that of \Excel but does
not use the calc chain. \Bazel is not minimal in the sense that it may restart a
task multiple times as new dependencies are discovered and rebuilt, but it
supports the early cutoff optimisation. Note that in practice the cost of
duplicate work due to the use of a restarting scheduler may often be just a
small fraction of the overall build cost~(\S\ref{sec-suspending}).

To support cloud builds, \Bazel maintains (i)~a \emph{content-addressable cache}
that maps the hash of a file's content to the actual content of that file;
(ii)~a memo table that records all executed build commands
with their input and output file hashes.
The memo table allows the build engine to bypass the execution
of a task, by predicting the hash of the result from the hashes of its
dependencies; then the content-addressable cache allows the engine
to download the result (if needed) based on the result hash. Further details
and a concrete implementation will be provided
in~\S\ref{sec-constructive-traces} and~\S\ref{sec-implementation-cloud}.

@@

\begin{table}
\vspace{2mm}
\caption{Build system differences.\label{tab-summary}}
\vspace{-1mm}
% \smaller
\centering
\begin{tabular}{llllccc}
\hline
Build system\gap & Persistent build information\gap & Scheduler\gap   & Dependencies\gap    & Minimal\gap & Cutoff\gap & Cloud\\\hline
\Make       & File modification times      & Topological\gap & Static          & Yes     & No     & No   \\
\Excel      & Dirty cells, calc chain      & Restarting  & Dynamic         & No      & No     & No   \\
\Shake      & Previous dependency graph    & Suspending  & Dynamic         & Yes     & Yes    & No   \\
\Bazel      & Cloud cache, command history\gap & Restarting  & Dynamic$^{(*)}$ & No      & Yes    & Yes  \\\hline
\end{tabular}
% \vspace{2mm}
$^{(*)}$At present, user-defined build rules cannot have dynamic dependencies.
\vspace{1mm}
\end{table}

### 2019-10-19 papers/jfp/3-abstractions.tex
Work on S1-3

\begin{minted}[xleftmargin=10pt]{text}
A1: 10     B1: A1 + A2
A2: 20     B2: B1 * 2
\end{minted}

@@

\noindent
We instantiate keys \hs{k} with \hs{String}, and values \hs{v} with \hs{Integer}.
(Real spreadsheet cells would contain a wider range of values, of course.) The
task description \hs{sprsh1} embodies all the \emph{formulae} of the spreadsheet,
but not the input values. It pattern-matches on the key to see if it has a task
description (in the \Excel case, a formula) for it. If not, it returns
\hs{Nothing}, indicating that the key is an input. If there is a formula in the
cell, it returns the \hs{Task} to compute the value of the formula. Every
\hs{Task} is given a \emph{callback} \hs{fetch} to find the value of any keys on
which it depends.

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Applicative functors
pure  :: Applicative f => a -> f a
(<$>) :: Functor     f =>   (a -> b) -> f a -> f b -- Left-associative
(<*>) :: Applicative f => f (a -> b) -> f a -> f b -- Left-associative
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard State monad from Control.Monad.State
data State s a
instance Monad (State s)
get       :: State s s
gets      :: (s -> a) -> State s a
put       :: s -> State s ()
modify    :: (s -> s) -> State s ()
runState  :: State s a -> s -> (a, s)
execState :: State s a -> s -> s
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard types from Data.Functor.Identity and Data.Functor.Const
newtype Identity a = Identity { runIdentity :: a }
newtype Const m a  = Const    { getConst    :: m }
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
instance Functor (Const m) where
    fmap _ (Const m) = Const m
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
instance Monoid m => Applicative (Const m) where
    pure _              = Const mempty   -- mempty is identity for monoid m
    Const x <*> Const y = Const (x <> y) -- <> is the binary operation for m
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard types from Control.Monad.Trans.Writer
newtype WriterT w m a = WriterT { runWriterT :: m (a, w) }
tell :: Monad m => w   -> WriterT w m () -- write a value to the log
lift :: Monad m => m a -> WriterT w m a  -- lift an action into WriterT
\end{minted}

% \vspace{-3mm}
\caption{Standard library definitions.}\label{fig-stdlib}
% \vspace{-4mm}
\end{figure}

@@

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
type Build c i k v = Tasks c k v -> k -> Store i k v -> Store i k v
\end{minted}
\vspace{1mm}

\noindent
The signature is very straightforward. Given a task description, a target key,
and a store, the build system returns a new store in which the value of the
target key is up to date. What exactly does ``up to date'' mean?  We answer
that precisely in \S\ref{sec-build-correctness}. Here is a simple build system:

@@

\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ store  = initialise () (\key -> if key == "A1" then 10 else 20)
@\ghci@ result = busy sprsh1 "B2" store
@\ghci@ getValue "B1" result
30
@\ghci@ getValue "B2" result
60
\end{minted}

>>>>

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{text}
A1: 10     B1: A1 + A2
A2: 20     B2: B1 * 2
\end{minted}
\vspace{1mm}

@@

\noindent
We instantiate keys \hs{k} with \hs{String}, and values \hs{v} with \hs{Integer}.
(Real spreadsheet cells would contain a wider range of values, of course.) The
task description \hs{sprsh1} embodies all the \emph{formulae} of the spreadsheet,
but not the input values. It pattern-matches on the key to see if it has a task
description (in the \Excel case, a formula) for it. If not, it returns
\hs{Nothing}, indicating that the key is an input. If there is a formula in the
cell, it returns the \hs{Task} to compute the value of the formula. Every
task is given a \emph{callback} \hs{fetch} to find the value of any keys on
which it depends. To run a \hs{Task}, we simply apply the function it holds to a
callback (see the definition of the function \hs{run} in Fig.~\ref{fig-types}).

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Applicative functors
pure  :: Applicative f => a -> f a
(<$>) :: Functor     f =>   (a -> b) -> f a -> f b -- Left-associative
(<*>) :: Applicative f => f (a -> b) -> f a -> f b -- Left-associative
\end{minted}
\vspace{0.5mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard State monad from Control.Monad.State
data State s a
instance Monad (State s)
get       :: State s s
gets      :: (s -> a) -> State s a
put       :: s -> State s ()
modify    :: (s -> s) -> State s ()
runState  :: State s a -> s -> (a, s)
execState :: State s a -> s -> s
\end{minted}
\vspace{0.5mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard types from Data.Functor.Identity and Data.Functor.Const
newtype Identity a = Identity { runIdentity :: a }
newtype Const m a  = Const    { getConst    :: m }
\end{minted}
\vspace{0.5mm}
\begin{minted}[fontsize=\small]{haskell}
instance Functor (Const m) where
    fmap _ (Const m) = Const m
\end{minted}
\vspace{0.5mm}
\begin{minted}[fontsize=\small]{haskell}
instance Monoid m => Applicative (Const m) where
    pure _              = Const mempty   -- mempty is identity for monoid m
    Const x <*> Const y = Const (x <> y) -- <> is the binary operation for m
\end{minted}
\vspace{0.5mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard types from Control.Monad.Trans.Writer
newtype WriterT w m a = WriterT { runWriterT :: m (a, w) }
tell :: Monad m => w   -> WriterT w m () -- write a value to the log
lift :: Monad m => m a -> WriterT w m a  -- lift an action into WriterT
\end{minted}
% \vspace{-3mm}
\caption{Standard library definitions.}\label{fig-stdlib}
% \vspace{-4mm}
\end{figure}

@@

\vspace{0.5mm}
\begin{minted}[xleftmargin=10pt]{haskell}
type Build c i k v = Tasks c k v -> k -> Store i k v -> Store i k v
\end{minted}
\vspace{0.5mm}

\noindent
The signature is very straightforward. Given a task description, a target key,
and a store, the build system returns a new store in which the value of the
target key is up to date. What exactly does ``up to date'' mean?  We answer
that precisely in \S\ref{sec-build-correctness}.

\noindent
Here is a simple build system:

@@

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ store  = initialise () (\key -> if key == "A1" then 10 else 20)
@\ghci@ result = busy sprsh1 "B2" store
@\ghci@ getValue "B1" result
30
@\ghci@ getValue "B2" result
60
\end{minted}
\vspace{1mm}
